---
title: "Trying TWD_norm on MPJ data"
output:
  html_document:
    df_print: paged
bibliography: references.bib
---

# Tasks

-   [ ] i need to figure out the growing season window and what would be a good approach to zeroing the TWD data because without that the TWDnorm reaches values much higher than 1 (some trees are in 1000+ range)

-   [ ] Something I need to think about is if TWD_predawn (min TWD before sunrise) is a good variable for TWD_norm so that TWD_norm is comparable to thermal indices.

-   [ ] Add citations like: [@peters2025]

# Summary

apply method from Normalized tree water deficit: an automated dendrometer signal to quantify drought stress in trees by Peters et al <https://doi.org/10.1111/nph.70266>

```{r libraries, include=FALSE}
library(tidyverse)
library(here)
library(plotly)
library(suncalc)
library(janitor)
library(patchwork)

source(here("code/supp/mpj_time_utils.R"))

```

```{r globals}

# MPJ site coords (edit if needed)
MPJ_LAT <- 34.43
MPJ_LONG <- -106.13

# To make naming axises consistent and easier

## TWD and MDS (µm)
LABEL_TWD <- labs(y = "TWD (µm)")
LABEL_TWD_PD <- labs(y = "TWD_predawn (µm)") # twd_predawn is minimum TWD value before dawn of that day
LABEL_MDS <- labs(y = "MDS (µm)") 

## Normalized TWD and MDS (Unitless)
LABEL_TWD_NORM <- labs(y = "TWD_norm (-)") 
LABEL_MDS_NORM <- labs(y = "MDS_norm (-)")

## Timestamps
LABEL_TS <- labs(x = "Time")
LABEL_TS_15MIN <- labs(x = "Time (15-min)")
LABEL_TS_DATE <- labs(x = "Date")


```

```{r load data, message=FALSE, warning=FALSE}

# read pj_dendro
twd_dat <- read_csv(here("data/processed/PJ_DENDRO.csv")) |>
  mutate(ts = relabel_mpj_wall_time(ts)) |>
  filter(!series %in% c("JCP292","JCP738","PCP728","PCP754")) |> 
  select(-"...1") |> 
  mutate(
    species = case_when(
      str_starts(series, "J") ~ "jumo",
      str_starts(series, "P") ~ "pied",
      TRUE ~ NA_character_
      )
    )

# precip for 2024-2025 should already be in data/processed/precip_24_25_daily.rds.
# uncommment the following script if you need to do it again, or add different years, but make sure you add the needed NMEG file to the data/raw folder.
#
source(here("code/supp/merge_precip_24_25.R"))

precip_24_25_daily <- read_rds(here("data/processed/precip_24_25_daily.rds")) 
  
summary(twd_dat)

summary(precip_24_25_daily)

```

# Quick visual

```{r }

p <- ggplot(twd_dat, aes(x = ts, y = twd, color = series)) +
  geom_line() + 
  LABEL_TWD +
  LABEL_TS_15MIN +
  facet_wrap(~species, ncol = 1)

ggplotly(p)

p <- ggplot(precip_24_25_daily, aes(x = date, y = precip_daily)) +
  geom_col() +
  LABEL_TS_DATE +
  labs(title = "Daily precipitation 2025", y = "total day precipitation (mm)")

ggplotly(p)
```

# Dendro

## Getting Dendrometer Stats

**MDS — Maximum Daily Shrinkage**

-   MDS = pre-dawn max(radius) - midday min(radius)

-   Following the workflow described by Peters et al., MDS is taken from
    the daily shrinkage cycle between the pre-dawn maximum stem radius
    and the minimum radius later in the day.

**TWD_predawn — Pre-dawn Tree Water Deficit**

-   TWD_predawn = daily minimum TWD before sunrise

-   This is the raw TWD term used in the Peters et al. normalization,
    which makes it the daily TWD statistic of interest here rather than
    midday TWD4.

**Normalized metrics**

$$
\text{Eqn 1: } TWD_{\text{predawn,norm}} = \frac{TWD_{\text{predawn}}}{MDS_{\max}}
$$

$$
\text{Eqn 2: } MDS_{\text{norm}} = \frac{MDS}{MDS_{\max}}
$$

-   \(MDS_{\max}\) is the 99th percentile of daily MDS for each
    individual dendrometer series over the available record.

### Prep derivation objects

Using `P2` as the worked example below because it has one of the
longest continuous MPJ records and makes the derivation plots easy to
follow.

```{r prep derivation objects}
example_series <- "P2"

dates <- seq.Date(min(as.Date(twd_dat$ts)), max(as.Date(twd_dat$ts)), by = "day")

sun_times <- getSunlightTimes(
  date = dates,
  lat = MPJ_LAT,
  lon = MPJ_LONG,
  keep = c("sunrise", "solarNoon", "sunset"),
  tz = MPJ_SOLAR_TZ
) |>
  mutate(date = as.Date(date)) |>
  select(date, sunrise, solarNoon, sunset)

twd_dat <- twd_dat |>
  mutate(date = as.Date(ts)) |>
  left_join(sun_times, by = "date") |>
  mutate(
    is_predawn = ts < sunrise,
    is_midday = ts >= solarNoon & ts <= sunset
  )

calc_daily_metrics <- function(data) {
  twd_daily <- data |>
    group_by(series, species, date) |>
    summarise(
      twd_predawn = if (any(is_predawn & !is.na(twd))) min(twd[is_predawn], na.rm = TRUE) else NA_real_,
      twd_predawn_ts = if (any(is_predawn & !is.na(twd))) {
        ts[is_predawn & !is.na(twd)][which.min(twd[is_predawn & !is.na(twd)])][1]
      } else {
        as.POSIXct(NA, tz = MPJ_FIXED_TZ)
      },
      .groups = "drop"
    )

  mds_daily <- data |>
    group_by(series, species, date) |>
    summarise(
      predawn_max = if (any(is_predawn & !is.na(value))) max(value[is_predawn], na.rm = TRUE) else NA_real_,
      predawn_max_ts = if (any(is_predawn & !is.na(value))) {
        ts[is_predawn & !is.na(value)][which.max(value[is_predawn & !is.na(value)])][1]
      } else {
        as.POSIXct(NA, tz = MPJ_FIXED_TZ)
      },
      midday_min = if (any(is_midday & !is.na(value))) min(value[is_midday], na.rm = TRUE) else NA_real_,
      midday_min_ts = if (any(is_midday & !is.na(value))) {
        ts[is_midday & !is.na(value)][which.min(value[is_midday & !is.na(value)])][1]
      } else {
        as.POSIXct(NA, tz = MPJ_FIXED_TZ)
      },
      mds = predawn_max - midday_min,
      .groups = "drop"
    )

  twd_daily |>
    left_join(mds_daily, by = c("series", "species", "date")) |>
    group_by(series) |>
    mutate(
      mds_max = if (all(is.na(mds))) {
        NA_real_
      } else {
        quantile(mds, probs = 0.99, na.rm = TRUE, names = FALSE, type = 7)
      },
      twd_norm = twd_predawn / mds_max,
      mds_norm = mds / mds_max,
      drought_stress_onset = twd_norm >= 1
    ) |>
    ungroup()
}

daily_metrics <- calc_daily_metrics(twd_dat)

example_dendro <- twd_dat |>
  filter(series == example_series) |>
  arrange(ts)

example_daily_metrics <- daily_metrics |>
  filter(series == example_series)

example_twd_predawn <- example_daily_metrics |>
  filter(!is.na(twd_predawn_ts)) |>
  transmute(ts = twd_predawn_ts, date, twd_predawn)

example_precip_daily <- precip_24_25_daily |>
  filter(
    date >= min(example_daily_metrics$date, na.rm = TRUE),
    date <= max(example_daily_metrics$date, na.rm = TRUE)
  )

p_rain_daily <- plot_ly(example_precip_daily) |>
  add_lines(
    x = ~date,
    y = ~precip_daily,
    name = "Daily precip",
    line = list(color = "blue")
  ) |>
  layout(
    yaxis = list(title = "Precip (mm)")
  )
```

### How are TWD and TWD_norm derived?

This mirrors the derivation style from the thermal-analysis notebook,
but uses `TWD_predawn` instead of `TWD4` so the walkthrough matches the
Peters et al. normalization workflow.

```{r derive twd norm, fig.width=12, fig.height=12}
p_rad_growth <- plot_ly(
  example_dendro,
  x = ~ts, y = ~value,
  type = "scatter", mode = "lines",
  name = "Stem radius",
  line = list(color = "forestgreen", width = 3)
) |>
  add_trace(
    x = ~ts,
    y = ~max,
    name = "Max radius envelope",
    mode = "lines",
    fill = "tonexty",
    fillcolor = "rgba(173, 216, 230, 0.35)",
    line = list(color = "firebrick", width = 3)
  ) |>
  layout(
    yaxis = list(title = "Stem Radius (µm)")
  )

p_twd <- plot_ly(example_dendro) |>
  add_lines(
    x = ~ts,
    y = ~twd,
    name = "TWD",
    line = list(color = "steelblue", width = 3)
  ) |>
  add_markers(
    data = example_twd_predawn,
    x = ~ts,
    y = ~twd_predawn,
    name = "TWD_predawn",
    marker = list(color = "black", size = 7, symbol = "circle")
  ) |>
  layout(
    yaxis = list(title = "TWD (µm)")
  )

p_twd_predawn_daily <- plot_ly(
  example_daily_metrics,
  x = ~date,
  y = ~twd_predawn,
  type = "scatter",
  mode = "lines",
  line = list(color = "black", width = 3),
  name = "TWD_predawn"
) |>
  layout(
    yaxis = list(title = "TWD_predawn (µm)"),
    xaxis = list(title = "Date")
  )

p_raw_daily <- plot_ly(example_daily_metrics) |>
  add_lines(
    x = ~date,
    y = ~twd_predawn,
    name = "TWD_predawn",
    line = list(color = "black", width = 3)
  ) |>
  add_lines(
    x = ~date,
    y = ~mds_max,
    name = "MDSmax (99th percentile)",
    line = list(color = "red", width = 2, dash = "dash")
  ) |>
  layout(
    yaxis = list(title = "Raw metric (µm)"),
    xaxis = list(title = "Date")
  )

p_twd_norm_example <- plot_ly(
  example_daily_metrics,
  x = ~date,
  y = ~twd_norm,
  type = "scatter",
  mode = "lines",
  line = list(color = "black", width = 3),
  name = "TWD_norm"
) |>
  add_lines(
    x = ~date,
    y = ~rep(1, nrow(example_daily_metrics)),
    name = "TWD_norm = 1",
    line = list(color = "red", width = 2, dash = "dash")
  ) |>
  layout(
    yaxis = list(title = "TWD_norm (-)"),
    xaxis = list(title = "Date")
  )

fig <- subplot(
  p_rad_growth,
  p_twd,
  p_twd_predawn_daily,
  p_raw_daily,
  p_twd_norm_example,
  p_rain_daily,
  nrows = 6,
  shareX = TRUE,
  titleY = TRUE
) |>
  layout(
    xaxis = list(title = "Time")
  )

fig
```

### How is MDS and MDS_norm derived?

The same daily windows can be reused for MDS: select the pre-dawn
maximum radius, select the midday minimum radius, take the difference,
and then divide by the same \(MDS_{\max}\) used above.

```{r derive mds norm, fig.width=12, fig.height=12}
p_mds_points <- plot_ly(example_dendro) |>
  add_lines(
    x = ~ts,
    y = ~value,
    name = "Stem radius",
    line = list(color = "forestgreen", width = 3)
  ) |>
  add_markers(
    data = example_daily_metrics,
    x = ~predawn_max_ts,
    y = ~predawn_max,
    marker = list(color = "red", size = 7),
    name = "Predawn max"
  ) |>
  add_markers(
    data = example_daily_metrics,
    x = ~midday_min_ts,
    y = ~midday_min,
    marker = list(color = "blue", size = 7),
    name = "Midday min"
  ) |>
  layout(
    yaxis = list(title = "Stem Radius (µm)")
  )

p_mds_daily <- plot_ly(
  example_daily_metrics,
  x = ~date,
  y = ~mds,
  type = "scatter",
  mode = "lines",
  line = list(color = "purple", width = 3),
  name = "MDS"
) |>
  add_lines(
    x = ~date,
    y = ~mds_max,
    name = "MDSmax (99th percentile)",
    line = list(color = "red", width = 2, dash = "dash")
  ) |>
  layout(
    yaxis = list(title = "MDS (µm)"),
    xaxis = list(title = "Date")
  )

p_mds_norm_example <- plot_ly(
  example_daily_metrics,
  x = ~date,
  y = ~mds_norm,
  type = "scatter",
  mode = "lines",
  line = list(color = "purple", width = 3),
  name = "MDS_norm"
) |>
  add_lines(
    x = ~date,
    y = ~rep(1, nrow(example_daily_metrics)),
    name = "MDS_norm = 1",
    line = list(color = "red", width = 2, dash = "dash")
  ) |>
  layout(
    yaxis = list(title = "MDS_norm (-)"),
    xaxis = list(title = "Date")
  )

fig <- subplot(
  p_mds_points,
  p_mds_daily,
  p_mds_norm_example,
  p_rain_daily,
  nrows = 4,
  shareX = TRUE,
  titleY = TRUE
) |>
  layout(
    xaxis = list(title = "Time")
  )

fig
```

### TWD_predawn and MDS together

```{r twd and mds together, fig.width=12, fig.height=12}
fig <- subplot(
  p_rad_growth,
  p_twd,
  p_twd_predawn_daily,
  p_mds_daily,
  p_twd_norm_example,
  p_mds_norm_example,
  p_rain_daily,
  nrows = 7,
  shareX = TRUE,
  titleY = TRUE
) |>
  layout(
    xaxis = list(title = "Time")
  )

fig
```

# Normalized Metrics

### Full timeseries across all dendrometers

```{r normalized full timeseries}
metrics_full <- daily_metrics |>
  filter(
    date >= min(daily_metrics$date, na.rm = TRUE),
    date <= max(daily_metrics$date, na.rm = TRUE)
  )

precip_full_daily <- precip_24_25_daily |>
  filter(
    date >= min(metrics_full$date, na.rm = TRUE),
    date <= max(metrics_full$date, na.rm = TRUE)
  )

p_twd_norm_all <- ggplot(metrics_full, aes(x = date, y = twd_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  LABEL_TWD_NORM +
  LABEL_TS_DATE +
  labs(title = "Normalized TWD for full timeseries")

ggplotly(p_twd_norm_all)

p_mds_norm_all <- ggplot(metrics_full, aes(x = date, y = mds_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  LABEL_MDS_NORM +
  LABEL_TS_DATE +
  labs(title = "Normalized MDS for full timeseries")

ggplotly(p_mds_norm_all)

p1 <- ggplot(metrics_full, aes(x = date, y = twd_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Normalized TWD (full record)", y = "TWD_norm (-)")

p2 <- ggplot(metrics_full, aes(x = date, y = mds_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Normalized MDS (full record)", y = "MDS_norm (-)")

p3 <- ggplot(precip_full_daily, aes(x = date, y = precip_daily)) +
  geom_col() +
  LABEL_TS_DATE +
  labs(y = "Daily precip. (mm)")

subplot(
  ggplotly(p1),
  ggplotly(p2),
  ggplotly(p3),
  nrows = 3,
  shareX = TRUE,
  heights = c(0.4, 0.4, 0.2),
  titleY = TRUE
)
```

**Remarks:**

I am still somewhat hesitant to over-interpret the full-record plots
because several dendrometer records begin after July 2024, which means
their estimate of \(MDS_{\max}\) is still based on an incomplete record.

The derivation figures above help show why large normalized values are
possible here: if pre-dawn TWD stays elevated while \(MDS_{\max}\)
remains relatively modest, then both `TWD_norm` and `MDS_norm` can stay
well above 1 for long periods.

### 2025-only zeroing

Trying the same workflow after zeroing each series at its first 2025
measurement. This is useful as a sensitivity test, but it removes the
multi-year framing of \(MDS_{\max}\).

```{r normalized 2025}
baseline_2025 <- twd_dat |>
  filter(date >= as.Date("2025-01-01")) |>
  group_by(series) |>
  slice_min(ts, n = 1, with_ties = FALSE) |>
  select(series, twd_zero = twd)

twd_dat_2025 <- twd_dat |>
  left_join(baseline_2025, by = "series") |>
  mutate(
    twd = twd - twd_zero
  ) |>
  filter(date >= as.Date("2025-01-01"))

daily_metrics_2025 <- calc_daily_metrics(twd_dat_2025)

precip_25_daily <- precip_24_25_daily |>
  filter(
    date >= min(daily_metrics_2025$date, na.rm = TRUE),
    date <= max(daily_metrics_2025$date, na.rm = TRUE)
  )

p_twd_predawn_2025 <- ggplot(daily_metrics_2025, aes(x = date, y = twd_predawn, color = series)) +
  geom_line() +
  facet_wrap(~species, ncol = 1) +
  LABEL_TWD_PD +
  LABEL_TS_DATE +
  labs(title = "Predawn TWD after 2025 zeroing")

ggplotly(p_twd_predawn_2025)

p_twd_norm_2025 <- ggplot(daily_metrics_2025, aes(x = date, y = twd_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  LABEL_TWD_NORM +
  LABEL_TS_DATE +
  labs(title = "Normalized TWD after 2025 zeroing")

ggplotly(p_twd_norm_2025)

p_mds_norm_2025 <- ggplot(daily_metrics_2025, aes(x = date, y = mds_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  LABEL_MDS_NORM +
  LABEL_TS_DATE +
  labs(title = "Normalized MDS after 2025 zeroing")

ggplotly(p_mds_norm_2025)

p1 <- ggplot(daily_metrics_2025, aes(x = date, y = twd_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Normalized TWD for 2025", y = "TWD_norm (-)")

p2 <- ggplot(daily_metrics_2025, aes(x = date, y = mds_norm, color = series)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Normalized MDS for 2025", y = "MDS_norm (-)")

p3 <- ggplot(precip_25_daily, aes(x = date, y = precip_daily)) +
  geom_col() +
  LABEL_TS_DATE +
  labs(y = "Daily precip. (mm)")

subplot(
  ggplotly(p1),
  ggplotly(p2),
  ggplotly(p3),
  nrows = 3,
  shareX = TRUE,
  heights = c(0.4, 0.4, 0.2),
  titleY = TRUE
)



```

[**Thoughts on the 2025-only view:**]{.underline}

This sensitivity check should make it easier to judge how much of the
very large normalized values are coming from the cumulative TWD baseline
versus the within-2025 stem-radius dynamics themselves. It also gives a
matched view of `TWD_norm` and `MDS_norm`, which was missing from the
earlier version of this notebook.

# Norm Species averaged

Here the species-level normalized metrics are calculated by averaging the
sub-daily dendrometer signal within each species first, then deriving
`TWD_predawn`, `MDS`, `TWD_norm`, and `MDS_norm` from that averaged
species time series.

```{r Norm Species averaged}

bad_tomsts <- c("P13","J21","J23")

twd_dat_species_mean <- twd_dat |>
  filter(!series %in% bad_tomsts) |> 
  group_by(species, ts, date, sunrise, solarNoon, sunset) |>
  summarise(
    value = mean(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(species, ts) |>
  group_by(species) |>
  mutate(
    series = species,
    max = cummax(value),
    twd = max - value,
    is_predawn = ts < sunrise,
    is_midday = ts >= solarNoon & ts <= sunset
  ) |>
  ungroup()

species_metrics_full <- calc_daily_metrics(twd_dat_species_mean)

precip_species_daily <- precip_24_25_daily |>
  filter(
    date >= min(species_metrics_full$date, na.rm = TRUE),
    date <= max(species_metrics_full$date, na.rm = TRUE)
  )

p_species_twd <- ggplot(twd_dat_species_mean, aes(x = ts, y = twd, color = species)) +
  geom_line() +
  LABEL_TWD +
  LABEL_TS_15MIN +
  labs(title = "Species-averaged TWD before daily metric derivation")

ggplotly(p_species_twd)

p_species_twd_norm <- ggplot(species_metrics_full, aes(x = date, y = twd_norm, color = species)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  LABEL_TWD_NORM +
  LABEL_TS_DATE +
  labs(title = "Species-averaged TWD_norm")

ggplotly(p_species_twd_norm)

p_species_mds_norm <- ggplot(species_metrics_full, aes(x = date, y = mds_norm, color = species)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  LABEL_MDS_NORM +
  LABEL_TS_DATE +
  labs(title = "Species-averaged MDS_norm")

ggplotly(p_species_mds_norm)

p1 <- ggplot(species_metrics_full, aes(x = date, y = twd_norm, color = species)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  labs(title = "Species-averaged TWD_norm", y = "TWD_norm (-)")

p2 <- ggplot(species_metrics_full, aes(x = date, y = mds_norm, color = species)) +
  geom_line() +
  geom_hline(yintercept = 1) +
  labs(title = "Species-averaged MDS_norm", y = "MDS_norm (-)")

p3 <- ggplot(precip_species_daily, aes(x = date, y = precip_daily)) +
  geom_col() +
  LABEL_TS_DATE +
  labs(y = "Daily precip. (mm)")

subplot(
  ggplotly(p1),
  ggplotly(p2),
  ggplotly(p3),
  nrows = 3,
  shareX = TRUE,
  heights = c(0.4, 0.4, 0.2),
  titleY = TRUE
)

```
